---
title: "Severe Weather Events Impacts on Both Public Health and Economic Problems for Communities and Municipalities"
output: 
  html_document:
    keep_md: true
---
# Synopsis
The report explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm data for the period  year 1950 and end in November 2011.
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data Processing
```{r, echo=TRUE}
library(plyr)
stormData <- read.csv("StormData.csv", header=TRUE)

names(stormData)

dim(stormData)
#(stormData$EVTYPE)

unique(stormData$PROPDMGEXP)
unique(stormData$CROPDMGEXP)
# head(stormData, 100)
smallData <- stormData[,c(8,23,24)]

head(smallData)
tail(smallData)

sortHelper <- function(fieldName, top = 15, dataset = stormData) {

    index <- which(colnames(dataset) == fieldName)

    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")

    names(field) <- c("EVTYPE", fieldName)

    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

fatalities <- sortHelper("FATALITIES", dataset = stormData)
fatalities

injuries <- sortHelper("INJURIES", dataset = stormData)
injuries
rm(stormData)
```

```{r, echo=TRUE, results='asis'}
stormData <- read.csv("StormData.csv", header=TRUE)

# $$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$

# Transform PROPDMG multiplier # Vertical subset - limit the data to the required fields.
#stormData <- stormData[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG","PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
# Sorting the property exponent data
stormData$PROPEXP[stormData$PROPDMGEXP == "K"] <- 1000
stormData$PROPEXP[stormData$PROPDMGEXP == "M"] <- 10^6
stormData$PROPEXP[stormData$PROPDMGEXP == ""] <- 1
stormData$PROPEXP[stormData$PROPDMGEXP == "B"] <- 10^9
stormData$PROPEXP[stormData$PROPDMGEXP == "m"] <- 10^6
stormData$PROPEXP[stormData$PROPDMGEXP == "0"] <- 1
stormData$PROPEXP[stormData$PROPDMGEXP == "5"] <- 10^5
stormData$PROPEXP[stormData$PROPDMGEXP == "6"] <- 10^6
stormData$PROPEXP[stormData$PROPDMGEXP == "4"] <- 10000
stormData$PROPEXP[stormData$PROPDMGEXP == "2"] <- 100
stormData$PROPEXP[stormData$PROPDMGEXP == "3"] <- 1000
stormData$PROPEXP[stormData$PROPDMGEXP == "h"] <- 100
stormData$PROPEXP[stormData$PROPDMGEXP == "7"] <- 10^7
stormData$PROPEXP[stormData$PROPDMGEXP == "H"] <- 100
stormData$PROPEXP[stormData$PROPDMGEXP == "1"] <- 10
stormData$PROPEXP[stormData$PROPDMGEXP == "8"] <- 10^8
# give 0 to invalid exponent data, so they not count in
stormData$PROPEXP[stormData$PROPDMGEXP == "+"] <- 0
stormData$PROPEXP[stormData$PROPDMGEXP == "-"] <- 0
stormData$PROPEXP[stormData$PROPDMGEXP == "?"] <- 0
# compute the property damage value
stormData$PROPDMGVAL <- stormData$PROPDMG * stormData$PROPEXP


# Sorting the property exponent data
stormData$CROPEXP[stormData$CROPDMGEXP == "M"]  <- 10^6
stormData$CROPEXP[stormData$CROPDMGEXP == "K"]  <- 1000
stormData$CROPEXP[stormData$CROPDMGEXP == "m"]  <- 10^6
stormData$CROPEXP[stormData$CROPDMGEXP == "B"]  <- 10^9
stormData$CROPEXP[stormData$CROPDMGEXP == "0"]  <- 1
stormData$CROPEXP[stormData$CROPDMGEXP == "k"] <- 1000
stormData$CROPEXP[stormData$CROPDMGEXP == "2"]  <- 100
stormData$CROPEXP[stormData$CROPDMGEXP == ""]   <- 1

# give 0 to invalid exponent data, so they not count in
stormData$CROPEXP[stormData$CROPDMGEXP == "?"] <- 0

# compute the crop damage value
stormData$CROPDMGVAL <- stormData$CROPDMG * stormData$CROPEXP

# aggregate the data by event
fatal   <- aggregate(FATALITIES ~ EVTYPE, data = stormData, FUN = sum)
injury  <- aggregate(INJURIES ~ EVTYPE, data = stormData, FUN = sum)
propdmg <- aggregate(PROPDMGVAL ~ EVTYPE, data = stormData, FUN = sum)
cropdmg <- aggregate(CROPDMGVAL ~ EVTYPE, data = stormData, FUN = sum)



# $$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$$
library(xtable)
#packages <- c("xtable")
#sapply(packages, require, character.only = TRUE, quietly = TRUE)

# Subset Data into the columns of interrest
healthData <- stormData[, c("EVTYPE", "INJURIES", "FATALITIES")]
head(healthData)


#healthData <- aggregate.data.frame(x = healthData[, c(2, 3)], by = list(healthData$EVTYPE), FUN = sum)
healthData <- aggregate(x = healthData[, c(2, 3)], by = list(healthData$EVTYPE), FUN = sum)

# Rename the first coluns
colnames(healthData)[1] <- "EVTYPE"

# Add a column for totals
healthData$TOTAL <- healthData$FATALITIES + healthData$INJURIES

healthData <- healthData[order(healthData$TOTAL, decreasing = TRUE), ]

head(healthData)

# Rename the columns to corresponding event types
row.names(healthData) <- healthData$EVTYPE

# Remove all rows with total of zero
healthData <- healthData[healthData$TOTAL != 0, ]

head(healthData)

healthTable <- xtable(healthData[1:10, c("INJURIES","FATALITIES")], caption = "Table 1. Top Ten Events",digit=c(0,0,0))

#print(healthTable, floating=FALSE, comment=F,type="html",hline.after=c(0,nrow(healthTable)))
print(healthTable,type="html")
#healthTable

# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
economyData <- stormData[, c("EVTYPE", "PROPDMG", "CROPDMG")]
head(economyData)

economyData <- aggregate(x = economyData[, c(2, 3)], by = list(economyData$EVTYPE), FUN = sum)
colnames(economyData)[1] <- "EVTYPE"
economyData$TOTAL <- economyData$PROPDMG + economyData$CROPDMG
economyData <- economyData[order(economyData$TOTAL, decreasing = TRUE), ]
row.names(economyData) <- economyData$EVTYPE
economyData <- economyData[economyData$TOTAL != 0, ]
head(economyData)

economyTable <- xtable(economyData[1:10, c("PROPDMG", "CROPDMG")], caption = "Table 2. crop damage",digit=c(0,0,0))
#names(economy.table) <- c("Property Damage", "Crop Damage")
print(economyTable, type = "html")

# %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

rm(stormData)
```


## Results
```{r, echo=TRUE}
# get top10 event with highest fatalities
fatal10 <- fatal[order(-fatal$FATALITIES), ][1:10, ]
fatal10

# get top10 event with highest injuries
injury10 <- injury[order(-injury$INJURIES), ][1:10, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)

barplot(fatal10$FATALITIES, las = 3, names.arg = fatal10$EVTYPE, main = "Weather Events With The Top 10 Highest Fatalities", 
    ylab = "number of fatalities", col = "red")
barplot(injury10$INJURIES, las = 3, names.arg = injury10$EVTYPE, main = "Weather Events With the Top 10 Highest Injuries", 
    ylab = "number of injuries", col = "red")


########################################################
# get top 10 events with highest property damage
propdmg10 <- propdmg[order(-propdmg$PROPDMGVAL), ][1:10, ]
# get top 10 events with highest crop damage
cropdmg10 <- cropdmg[order(-cropdmg$CROPDMGVAL), ][1:10, ]
par(mfrow = c(1, 2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(propdmg10$PROPDMGVAL/(10^9), las = 3, names.arg = propdmg10$EVTYPE, 
    main = "Top 10 Events with Greatest Property Damages", ylab = "Cost of damages ($ billions)", 
    col = "red")
barplot(cropdmg10$CROPDMGVAL/(10^9), las = 3, names.arg = cropdmg10$EVTYPE, 
    main = "Top 10 Events With Greatest Crop Damages", ylab = "Cost of damages ($ billions)", 
    col = "red")
```