---
title: "Severe Weather Events Impacts on Both Public Health and Economic Problems for Communities and Municipalities"
output: 
  html_document:
    keep_md: true
  pdf_document:
    highlight: zenburn 
---
# Synopsis
The report explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm data for the period  year 1950 and end in November 2011.
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data Processing
```{r, echo=TRUE}
library(plyr)
stormData <- read.csv("StormData.csv", header=TRUE)

names(stormData)

dim(stormData)
#(stormData$EVTYPE)

unique(stormData$PROPDMGEXP)
unique(stormData$CROPDMGEXP)
# head(stormData, 100)
smallData <- stormData[,c(8,23,24)]

head(smallData)
tail(smallData)

sortHelper <- function(fieldName, top = 15, dataset = stormData) {

    index <- which(colnames(dataset) == fieldName)

    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")

    names(field) <- c("EVTYPE", fieldName)

    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

fatalities <- sortHelper("FATALITIES", dataset = stormData)
fatalities

injuries <- sortHelper("INJURIES", dataset = stormData)
injuries
```

```{r, echo=TRUE}

library(xtable)
packages <- c("xtable")
sapply(packages, require, character.only = TRUE, quietly = TRUE)

# Subset Data into the columns of interrest
healthData <- stormData[, c("EVTYPE", "INJURIES", "FATALITIES")]
#head(healthData)


#healthData <- aggregate.data.frame(x = healthData[, c(2, 3)], by = list(healthData$EVTYPE), FUN = sum)
healthData <- aggregate(x = healthData[, c(2, 3)], by = list(healthData$EVTYPE), FUN = sum)

# Rename the first coluns
colnames(healthData)[1] <- "EVTYPE"

# Add a column for totals
healthData$TOTAL <- healthData$FATALITIES + healthData$INJURIES

healthData <- healthData[order(healthData$TOTAL, decreasing = TRUE), ]

head(healthData)

# Rename the columns to corresponding event types
row.names(healthData) <- healthData$EVTYPE

# Remove all rows with total of zero
healthData <- healthData[healthData$TOTAL != 0, ]

head(healthData)

healthTable <- xtable(healthData[1:10, c("INJURIES","FATALITIES")], caption = "Table 1. Top Ten Events")

print(healthTable, floating=FALSE, comment=F,type="html",hline.after=c(0,nrow(healthTable)))
#healthTable

# http://rstudio-pubs-static.s3.amazonaws.com/18170_65d84748b47843c6a01d81b60b09d10c.html
# http://rstudio-pubs-static.s3.amazonaws.com/20314_2cc26c06d7334717acc71dae3bc4c05e.html
# http://rpubs.com/proudindiv/18069
# http://rstudio-pubs-static.s3.amazonaws.com/18276_dbaca171292d46088a5c766d548938f6.html
# http://rstudio-pubs-static.s3.amazonaws.com/25857_77d0955a90b24a4d80e15c6bdca441f9.html
# 
#####################################
### http://rstudio-pubs-static.s3.amazonaws.com/18170_65d84748b47843c6a01d81b60b09d10c.html
# http://www.inside-r.org/packages/cran/xtable/docs/print.xtable
```
## Results
